---
title: "A Simple Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A_Simple_Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# 1. Introduction:

`Coxtp` is an R package for fitting penalized Newton's method for the time-varying effects model using mAIC, TIC, GIC as information criteria, in particular we span the parameter using basis functions. Utilities for carrying out post-fitting visualization, summarize, and inference are also provided.


# 2. Installation:

```{r setup,include=TRUE}
library(surtvep)
```


# 3. Dataset preperation:


For the purpose of demonstration, we will use the simulated dataset "sim_data" in the our package. 

```{r}
sim_data=sim_data
```


Let’s check the data first:

```{r}
head(sim_data)
```
Then, let's extract the time and event as vector, and get the remaining information in the dataset as matrix

```{r}

event=sim_data[,"event"]
time=sim_data[,"time"]
data=sim_data[,!colnames(sim_data) %in% c("event","time")]


```

# 4. Model fitting

## 4.1 Newton Method without penalization:

### 4.1.1 Simple fitting:

Let's fit the model. Here, the default method is Newton Method without penalization, with smooth-spline. Term `lambda_spline` refers to the smoothing parameter lambda(Detail could be found under both "Model parameter(Add link)" section or our paper here(add link). Default value is 0, which refers to no penalization. 

```{r}

fit <- coxtp(event = event, z = data, time = time)

```
To get the estimated time-varying effect of a specific coefficient, we could use the following plot function in our package:

```{r}
coxtp.plot(fit,coef="V1")
```

The plot shows the result of the time-varying effect of Variable "V1".

### 4.1.2 Detailed calculations and theories.

#### 4.1.2.1 Model results meanings:

First, Let's look at the `fit` result

```{r}
summary(fit)
```
There are 4 results saved under the fit result. 

* `model_result` save the detailed model results which we will explain in a minute. 

* `lambda.selected` saved the best lambda chosen based on different criteria which will not be used for the non-penalized model. 

* `p` refers to the number of covariates used in the model 

* `z_names` records the covariates names. 


The detailed model result was saved in `model_result`, which can be called by `fit$model_result`. Now, let's explore the result a little bit:

```{r}
summary(fit$model_result)
```
Here we noticed that there are 18 items in the model results list. Following is an explanation of each item: 

* `theta`: Estimation matrix of $\theta$
* `logplkd`: log-partial likelihood 
* `theta_all`: Internal validation use
* `theta_list`: The estimation matrix at each Newton's update
* `AIC_all`: Akakia information criterion
* `TIC_all`: Takeuchi information criterion
* `TIC2_all`: Internal validation use
* `GIC_all`: Generalized information criterion
* `AIC_trace`: Internal validation use
* `TIC_trace`: Internal validation use
* `TIC2_trace`: Internal validation use
* `GIC_trace`: Internal validation use
* `logplkd_vec`: log-partial likelihood at each iteration
* `SplineType`: spline used for fitting the model
* `VarianceMatrix`: Variance Matrix
* `uniqfailtimes`: The input unique event times if `ties="Breslow"`, input time points if `ties="None"` (Add link)
* `bases`: The basis function used for estimating time-varying effects
* `knots`: Number of basis functions used for estimating time-varying effects

#### 4.1.2.2: How to get the effect of a specific time point?

We are more interested in estimating time-varying effect of the covariates. Following is an simple tutorial of how to do that. First, a little background about the time-varying effect in cox model(You could also check this part at our paper which have more detailed explanation.(insert link))

If we Let $X_i=(X_{i1},X_{i2},...X_{ip})^T$ refers to the $i_{th}$ individuals in the dataset with p covariates(which could also be understand as the $i_{th}$ row in the data we extract above). Let $\lambda(t|X_i)$ denote the hazard of having the event at time t for the $i_{th}$ individual, $\lambda_0(t)$ denote the hazard of having the event at time 0. When we considering the covarites having time fixed effect, we have the following formula for $\lambda(t|X_i)$:

$\lambda(t|X_i)=\lambda_0(t)exp(X_i^T\beta)$

Where $\beta$ refers to the coefficients where $\beta=(\beta_1,\beta_2,...\beta_p)$, which in this example, is (V1, V2). Which have similar format as the GLM model.
For time-varying effect model, we are simply replace $\beta$ with a set of $\beta(t)$. Thus, the time varying equations could be transferred as following:

$\lambda(t|X_i)=\lambda_0(t)exp(X_i^T\beta(t))$

Similar, $\beta(t)=(\beta_1(t),\beta_2(t),...\beta_p(t))$ where $\beta(.)$ refers to a set of cubic B-spline(Details for B-spline refers to here : insert link). Where, the single $\beta_p(t)$ could be estimated using the following formula:

$\beta_p(t)=\theta_p^TB(t)=\sum_{k=1}^K\theta_{pk}B_k(t)$

Here, K refers to the given number of knots.

Thus, to calculate the time varying effect of coefficient p, we just need to get both estimated B spline and the $\theta$ matrix. The B-spline was saved in `model_result$bases` and $\theta$ matrix was saved in the last item in  `model_result$theta_list` Following is the code for calculation:

```{r}
model_result  = fit$model_result
B.spline      = as.matrix(model_result$bases)
theta         = model_result$theta_list[[length(model_result$theta_list)]]

beta          = B.spline %*% t(theta)

dim(beta)
head(beta)


```

As a result, $\beta$ is a 2487*2 matrix(we are using the "Breslow" ties, thus there are 2487 rows instead of 5000, detail about ties and Breslow ties could refers to here(insert link)). We could also get the 95%CI for the estimation, which is calculates as below:

```{r}
B.spline <- as.matrix(model_result$bases)  
theta_plot  <- model_result$theta_list[[length(model_result$theta_list)]]

beta     <- B.spline%*%t(theta_plot)
var      <-  model_result$VarianceMatrix

colnames(beta)=fit$z_names
p        <- fit$p
knot     <- dim(B.spline)[2]
list     <- 1:dim(B.spline)[1]
beta_low <- matrix(0,dim(B.spline)[1],p)
beta_up  <- matrix(0,dim(B.spline)[1],p)
for(i in 1:p){
  beta_t_1   <- beta[,i]
  var2       <- var[((i-1)*knot+1):((i-1)*knot+knot),((i-1)*knot+1):((i-1)*knot+knot)]
  temp       <- 1.96*sqrt(vapply(list, function(x) matrix(B.spline[x,],1,knot)%*%var2%*%t(matrix(B.spline[x,],1,knot)),FUN.VALUE=numeric(1)))
  low        <- beta_t_1-temp
  up         <- beta_t_1+temp
  
  beta_low[,i] <- low
  beta_up[,i]  <- up
}

colnames(beta_low) <- paste0(fit$z_names,"_low")
colnames(beta_up) <- paste0(fit$z_names,"_up")

head(beta_low)
head(beta_up)
```

matrix `beta_low` records all the lower bound of beta and `beta_up` records all the upper bound of beta.

As a result, we could plot the effect by ourselves:

```{r}
beta              <- as.data.frame(beta)
beta              <- cbind(beta, beta_low, beta_up)
y=beta$V1
ymin=beta$V1_low
ymax=beta$V1_up
time=model_result$uniqfailtimes


ggplot(data=beta, aes(x=time)) + 
  geom_line(aes(y= y),size = 0.9,color = 'red') + 
  geom_ribbon(aes(ymin = ymin, ymax = ymax), fill="red", alpha = 0.2) + 
  scale_y_continuous(name='Hazard Ratio (log-scale)') +
  theme_bw() +  theme(plot.title = element_text(hjust = 0.5)) +
  theme(panel.grid.major = element_blank(),panel.grid.minor = element_blank()) +
  theme(text= element_text(size=14)) + theme(axis.text= element_text(size=14)) +
  theme(axis.title.y = element_text(margin= margin(t=0, r=10, b=0, l=0))) +labs(x="Time") + 
  ggtitle(paste0("Effect of V1 When holding other covariates constant"))
```



## 4.2 Newton Method with penalization:

We use the smooth spline here for penalization(default). You could also use `spline="P-spline"`. For `lambda_spline`, you could either enter an numeric number(must be integer) or a vector of numbers. If a vector of numbers were entered, the best lambda was selected based on different criteria(AIC, TIC or GIC). If `lambda_spline` was entered as a single number, the model result format was kind of similar as the 4.1, the model result could be called by `model_result`. 

Here we are using the `lambda_spline` following is a model fit with the `lambda_spline` as a vector for different illustration purpose:

```{r}
time=sim_data[,"time"]
lambda_spline_all=c(1,10,100,500,1000)
fit_penalized <- coxtp(event = event, z = data, time = time,lambda_spline=lambda_spline_all)

```
The optimal lambda was saved in the model term `lambda.selected`

```{r}
best_lambda=fit_penalized$lambda.selected
best_lambda
```
From the result above, we noticed that with different selection criteria, the best lambda selected is quite different. Here, we use the AIC criteria which set `lambda=1000`. The result for this model was saved in related criteria model and could be called as below:

```{r}
AIC_model  = fit_penalized$model.AIC


summary(AIC_model)
```
We could also directly draw the plot using the following command:

```{r}
coxtp.plot(fit_penalized,IC="AIC",coef="V1")
```

Compare with the plot1, we could see that the effect of "V1" was shrink to roughly linear here. 

The detail calculation of B-spline matrix, $\theta$ matrix and $\beta$ matrix was similar as that illustrate as 4.1.2.2.



